---
title: "Module 1.2"
subtitle: "Working With APIs"
format: html
execute: 
  echo: true
  message: false
  warning: false
---

:::{callout-tip}
## Prework

- Install the devtools package. Type `install.packages("devtools")` in your console. You will need this to install the `vdemdata` package becaue it is not on the CRAN Network. 

- Install the `vdemdta` package from GitHub. Type `devtools::install_github("vdeminstitute/vdemdata")` in your console. 
:::

## Overview 


## Wrangle WB data

Picking up where we left off... but let's add one more variable...and use lubridate to specify the most recent date. 

```{r}
#| label: wb_data

library(wbstats) 
library(lubridate)
library(dplyr)
library(janitor)

indicators <- c(flfp = "SL.TLF.CACT.FE.ZS", women_rep = "SG.GEN.PARL.ZS")

women_emp <- wb_data(indicators, start_date = 1974, end_date = year(Sys.Date())) |> 
  select(!iso2c) |> 
  rename(year = date) |> 
    mutate(
    flfp = round_to_fraction(flfp, denominator = 100), 
    women_rep = round_to_fraction(women_rep, denominator = 100)
  )

glimpse(women_emp)
```

## Filter observations, select and create new variables 

```{r}
#| label: democracy

library(vdemdata)
library(tibble)

democracy <- vdem |> 
  filter(year >= 1974)  |> 
  select(
    country = country_name,
    vdem_ctry_id = country_id, 
    year, 
    polyarchy = v2x_polyarchy, 
    gdp_pc = e_gdppc, 
    region = e_regionpol_6C
    ) |>
    mutate(
    region = recode(region, `1` = "Eastern Europe", 
                    `2` = "Latin America", 
                    `3` = "Middle East", 
                    `4` = "Africa", 
                    `5` = "The West", 
                    `6` = "Asia")
  ) |> 
  as_tibble()

glimpse(democracy)
```
## Add country code to democracy tibble

```{r}
library(countrycode)

democracy <- democracy |> 
  mutate(iso3c = countrycode(vdem_ctry_id, "vdem", "wb"))  |> 
  relocate(iso3c, .after = vdem_ctry_id)

glimpse(democracy)
#filter(democracy, is.na(iso3c))
```

## Merge two datasets 

```{r}
library(readr)

dem_women <- left_join(democracy, women_emp, by = c("iso3c", "year")) |> 
  rename(country = country.x) |> 
  select(!country.y)

glimpse(dem_women)  

write_csv(dem_women, "dem_women.csv")
```

## Group, summarize and arrange (summarize by region)

```{r}
dem_summary <- dem_women |> 
  group_by(region)  |> 
  summarize(
    polyarchy = mean(polyarchy, na.rm = TRUE),
    gdp_pc = mean(gdp_pc, na.rm = TRUE), 
    flfp = mean(flfp, na.rm = TRUE), 
    women_rep = mean(women_rep, na.rm = TRUE)
  ) |> 
  arrange(desc(polyarchy)) 

dem_summary

#filter(dem_women, is.na(region))
```
