---
title: "Module 2.2"
subtitle: "Accessibility, Clarity and Interactivity"
format: html
highlight-style: atom-one
execute:
  echo: true
  message: false
  warning: false
---

::: {.callout-tip}
## Prework

- Install `plotly` (`install.packages("plotly")`) and have a look at the [documentation](https://plotly.com/ggplot2/getting-started/#plotly-for-r)
- Install `colorBlindness` (`install.packages("colorBlindness")`) and read this [vignette](https://cran.r-project.org/web/packages/colorBlindness/vignettes/colorBlindness.html)
- Install `ggthemes` (`install.packages("ggthemes")`)
- Generate a quarto document named "module-2.2.qmd" in your "modules" project folder so that you can code along with me
- In your quarto document, run these code chunks and familiarize yourself the data frames that they generate 

```{r}
#| label: setup

library(readr)
library(dplyr)
library(ggplot2)
library(wbstats)
library(countrycode)

indicators = c(flfp = "SL.TLF.CACT.FE.ZS", gdp_pc = "NY.GDP.PCAP.KD") # define indicators

## Regional levels of FLFP for column chart 

flfp_gdp_regions <- 
  wb_data("SL.TLF.CACT.FE.ZS", country = "regions_only", mrnev = 1) |> 
  rename(
    region = country,
    year = date,
    flfp = SL.TLF.CACT.FE.ZS
  ) |> 
  select(region, iso3c, year, flfp)
  
flfp_gdp_regions

## Cross-section of data on FLFP and GDP per capita for scatter plot

flfp_gdp <- wb_data(indicators, mrnev = 1) |> # download most recent non-missing value of indicators
    left_join(select(wb_countries(), c(iso3c, region)), by = "iso3c")  # add regions from countrycode()

glimpse(flfp_gdp)

## Time series data on regional trends in FLFP for line chart

flfp_ts <- wb_data("SL.TLF.CACT.FE.ZS", country = "regions_only", start_date = 1990, end_date = 2022) |> 
  rename(
    region = country,
    year = date,
    flfp = SL.TLF.CACT.FE.ZS
  ) |> 
  select(region, iso3c, year, flfp)
  
glimpse(flfp_ts)
```
:::

## Overview

In this lesson we are going to take our visualizations from Module 2.1 and improve them. In the first part of the lesson we are going to focused on how to make your visualizations accessible to a color-blind audience. Then we will discuss how to improve the look of your visualizations with themes and to provide additional information and context with annotations. Finally, we will spend some time talking about how to make your graphs interactive so that users can explore them in a more dynamic and flexible way. 

## Color Schemes

{{< video https://www.youtube.com/embed/wo9vZccmqwc title = 'Accessible Color Schemes'>}}

There are a number of different [types](https://www.nei.nih.gov/learn-about-eye-health/eye-conditions-and-diseases/color-blindness/types-color-blindness) of color blindness, but the most common type is red-green color blindness. Making your visualizations colorblind-accessible can be important for convincing [certain audiences](https://towardsdatascience.com/is-your-color-palette-stopping-you-from-reaching-your-goals-bf3b32d2ac49). Most notably, approximately 8% of men are affected by color blindness. 

Let's start off by looking at the line chart that we did in the last module pertaining to Huntington's three waves of democratization:

```{r}
#| label: dem_waves
dem_waves_ctrs <- read_csv("data/dem_waves_ctrs.csv")

dem_waves_chart <- ggplot(dem_waves_ctrs, aes(x = year, y = polyarchy, color = country)) +
  geom_line(linewidth = 1) + 
  labs(
    x = "Year", 
    y = "Polyarchy Score", 
    title = 'Democracy in countries representing three different "waves"', 
    caption = "Source: V-Dem Institute", 
    color = "Country"
  )

dem_waves_chart
```

The problem with this plot is that people with red-green color blindness will not be able to distinguish between the lines for Japan and Portungal. There are many tools that we could use to see how this is true, but here we are going to focus on the color vision deficiency (CVD) simulator from the [colorBlindness](https://cran.r-project.org/web/packages/colorBlindness/vignettes/colorBlindness.html) package. To use it, all we have to do is load `colorBlindness` and call `cvdPlot()` on the stored plot. 

```{r}
#| label: dem_waves_cvd
#| fig-width: 12
#| fig-height: 7
library(colorBlindness)

cvdPlot(dem_waves_chart)
```

In this output, we can see how deuteranopia and protonopia (red-green color blindness) would experience our plot. Notice how hard it is to distinguish between Japan and Portugal here. We can also see how someone with monochromatic vision would see the plot by looking at the "desaturated (BW)" plot. While monochromatic vision is very rare, we can use this plot to make adjustments for a worst-case scenario. 

So once you determine that your color scheme is not colorblind friendly, what should you do? There are two basic solutions available to you: making your own colorblind friendly color scheme or use a package that produces colorblind-friendly schemes for you. 

### Create your own colorblind-friendly color scheme

The first way to make your plot more accessible is to create your own [discrete scale](https://ggplot2.tidyverse.org/reference/scale_manual.html) using `scale_fill_manual()` or `scale_color_manual()` (depending on what type of plot you are trying to draw). 

Let's try using an [example](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/) color scheme from the R Cookbook to fill in the bars of a column chart. Here we will make use of the regional levels of female labor force participation data that we prepped in the prework section of this module. First we create a vector of colors called `cb_palette`. Then we build our bar chart and save it as an object called `flfp_region`. Crucially, when we perform our ggplot call, we include `fill = region` as a third dimension in our `aes()` function. This is in addition to including the region code `iso3c` on the x-axis. Finally, we use our palette to shade our bars using `scale_fill_manual()`. 

```{r}
#| label: flfp_region
cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

flfp_region <- ggplot(flfp_gdp_regions, aes(x = reorder(iso3c, -flfp), y = flfp, fill = region)) +
  geom_col() + 
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    x = "Region", 
    y = "Avg. Female Labor Force Participation", 
    title = "Levels of female labor force participation by region", 
    fill = "Region",
    caption = "Source: World Bank"
    ) 

flfp_region + scale_fill_manual(values = cb_palette)
```

Now we can check our visualization using the `cvdPlot()` function from the `colorBlindness` package. 

```{r}
#| label: flfp_region_cvd
#| fig-width: 12
#| fig-height: 7
flfp_region <- flfp_region + scale_fill_manual(values = cb_palette)

cvdPlot(flfp_region)
```

Although it would still be difficult for a person with red-green color blindness to see some of the colors in the chart (like the pink of SSF or the bright green of LAC), it would still be possible for them to distinguish the colors of the bars based on what would appear to them as shades of grey. 

### Use Viridis

Another option is to use a package designed with accessibility issues in mind. One of the more popular ones is the [viridis()](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) package. To use it just load `viridis` and add a `viridis` color scheme to the plot. You have the option of choosing continuous, discrete or binned color schemes but here we use the discrete option `scale_fill_viridis_d()` because our regions constitute a discrete variable. You can also try different `viridis` color maps by inserting the various available schemes in the function. For example you can get the "plasma" map with `scale_fill_viridis_d(option = "plasma"). The default map is "viridis." 

```{r}
#| label: viridis
flfp_region + scale_fill_viridis_d()
```

Now let's run our bar chart with the `viridis` color mapping through `cvdPlot`: 

```{r}
#| label: viridis_cvd
#| fig-width: 12
#| fig-height: 7
flfp_region <- flfp_region + scale_fill_viridis_d()

cvdPlot(flfp_region)
```

This is pretty good. In fact it looks fairly close to the original. It is even to easy to see the differences in colors when looking at the desaturated output. Of course the first column for the black and white plot is difficult to distinguish from the plot's background, but we will learn how to change this up later in this module. 

### Use R color brewer

Another package available to us is the [RColorBrewer](https://cran.r-project.org/web/packages/RColorBrewer/index.html) package. Technically, `RColorBrewer` is a separate package but its color mappings come [available](https://ggplot2.tidyverse.org/reference/scale_brewer.html) as part of `ggplot2`, so we don't have to load `RColorBrewer` in order to use it with `ggplot2`. We can also use the color brewer [palette selector tool](https://colorbrewer2.org/) to help select palettes that are "colorblind safe." 

Let's try updating our column chart with a ColorBrewer color scheme:

```{r}
#| label: brewer
flfp_region + scale_fill_brewer(palette = "YlGn") 
```

And now lets run it through the `cvdPlot()` function. 

```{r}
#| label: brewer_cvd
#| fig-width: 12
#| fig-height: 7
flfp_region <- flfp_region + scale_fill_brewer(palette = "YlGn")

cvdPlot(flfp_region)
```

### Scaling for scatter plots and line charts

One thing to note is that when we want to adjust the color scheme for a scatter plot or line chart, we should use "scale_color" instead of "scale_fill", e.g. `scale_color_manual()`, `scale_color_viridis_d()`, `scale_color_brewer`. Let's try a couple of examples. First, let's add a `viridis` color map to a scatter plot. We will use the flfp_gdp data we prepped in the prework section to generate our scatter plot.

```{r}
#| label: scale_color_viridis()  
wealth_flfp <- ggplot(flfp_gdp, aes(x = gdp_pc, y = flfp)) + 
  geom_point(aes(color = region)) + # color points by region
  geom_smooth(method = "loess", linewidth = 1) +  # make the line a loess curve
  scale_x_log10(labels = scales::label_dollar()) + # stretch axis, add '$' format
  scale_y_continuous(labels = scales::label_percent(scale = 1)) + # add % label
  labs(
    x= "GDP per Capita", # x-axis title
    y = "Female Labor Force Participation", # y-axis title
    title = "Wealth and female labor force participation", # plot title
    caption = "Source: World Bank Development Indicators", # caption
    color = "Region" # legend title
    )

wealth_flfp + scale_color_viridis_d(option = "plasma")
```

Now let's try adding a ColorBrewer scheme to a line chart. Here we will use the `flfp_ts` data frame that we prepped in our prework routine.

```{r}
#| label: scale_color_brewer()
flfp_line <- ggplot(flfp_ts, aes(x = year, y = flfp, color = region)) +
  geom_line(linewidth = 1) + 
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    x = "Year", 
    y = "Female Labor Force Participation", 
    title = "Regional trends in female labor force participation", 
    caption = "Source: V-Dem Institute", 
    color = "Country"
  )

flfp_line + scale_color_brewer(palette = "YlOrRd")
```

## Themes

{{< video https://www.youtube.com/embed/wo9vZccmqwc >}}

Another thing that we can do to improve the overall look of our plots is to change the theme. Here is a list of [themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) that are available with `ggplot2`. 

There are also many [extension packages](https://towardsdatascience.com/themes-to-spice-up-visualizations-with-ggplot2-3e275038dafa) that you can use to apply even more themes, some of which we may encounter later in the course. 

For now, let's take a couple of plots that we developed earlier in the lesson and apply some `ggplot2` themes to them. We can do this by simply adding the name of the theme to our code. 

```{r}
#| label: theme_dark()
wealth_flfp + scale_color_viridis_d(option = "plasma") + theme_dark()
```

```{r}
#| label: theme_minimal()
flfp_region + scale_fill_brewer(palette = "YlGn") + theme_minimal()
```


## Annotations

{{< video https://www.youtube.com/embed/wo9vZccmqwc >}}

Sometimes it makes sense to include annotations in our charts. We can achieve this by applying the annotate() function. To add a text annotation, we include "text" for the first argument, then the value of x and y at which we want our annotation to appear, and finally the text of the annotation that we want to display. Let's try adding text to our indicating where high-, middle- and low-income countries are concentraed on the `wealth_flfp` scatter plot that we developed earlier. 

```{r}
#| label: annotate1
wealth_flfp <- wealth_flfp + scale_color_viridis_d(option = "plasma") + theme_minimal()

wealth_flfp + annotate("text", x = 90000, y = 75, label = "Wealthy") + 
  annotate("text", x = 1000, y = 80, label = "Low income") +
  annotate("text", x = 10000, y = 20, label = "Middle income")
```

Another common annotation involves combining text with a horizontal or vertical reference line. For a horizontal intercept line we include an additional geom called `geom_hline`. The first argument is the value `yintercept` where we want the reference line to cross. Then we can add additional arguments to define the style, color and size of the line. 

```{r}
#| label: annotate2
flfp_line <- flfp_line + scale_color_viridis_d() 

flfp_line + geom_hline(yintercept=52, linetype="dashed", color = "red", size = 1) +
  annotate("text", x = 1995, y = 55, label = "Global average")
```

The same logic applies for a vertical reference line, except this time the geom is called `geom_vline` and the first argument, `xintercept`, is the point at which we want the line to cross the x-axis. 

```{r}
#| label: annotate3
flfp_line <- flfp_line + scale_color_viridis_d() 

flfp_line + geom_vline(xintercept=2020, linetype = "dashed", size = 1) +
  annotate("text", x = 2017, y = 35, label = "Pandemic")
```


## Interactivity

{{< video https://www.youtube.com/embed/wo9vZccmqwc >}}

One final thing we can do, which is really fun, is to add interactivity to our plots with [plotly](https://plotly.com/r/). We can do this by simply calling 
[ggplotly()](https://plotly.com/ggplot2/) on our plot object. 

```{r}
#| label: plotly
library(plotly)

ggplotly(flfp_line)
```

In a lot of cases, we may want to control the tool tip of the plot. The tool tip is what appears when the user hovers over information on the chart. In this next example, the chart does not look that great unless we add the `tooltip` argument. But it is fairly simple to do. We just add the elements that we want to appear in a combine function, e.g. `c()`. In this case we will include region and female labor force participation in the tool tip.  

```{r}
#| label: tooltip
flfp_region <- flfp_region + scale_fill_brewer(palette = "YlGn") + theme_minimal()

ggplotly(flfp_region, tooltip = c("region", "flfp")) # controlling the tooltip output
```

Another thing we may want to do is to include some additional annotations. We might also notice that some of the things we include in the labs argument in `ggplot2` do not get picked up by `plotly` and that we have to add them back with a `layout(annotations = )` call. One additional idiosyncrasy is that any item that we want to include in the `plotly` chart has to be passed as an argument in the `ggplot` code. For example, we need to include `aes(label = country)` to view country in tool tip. But `plotly` does support the R native pipe operator and that makes it a little easier to layer on multiple annotations.  

```{r}
#| label: plotly_scatterplot
library(plotly)

wealth_flfp_plotly <- wealth_flfp  + 
  scale_color_viridis_d(option = "plasma") +
  theme_minimal() +
  aes(label = country)  # need so ggplot retains label for plotly

ggplotly(wealth_flfp_plotly, tooltip = c("country", "flfp", "gdp_pc")) |> 
  
  layout(annotations = list(text = "Source: World Bank Development Indicators",  
                            font = list(size = 10), showarrow = FALSE,
                            xref = 'paper', x = 1.1, xanchor = 'right', xshift = 0,
                            yref = 'paper', y = -.1, yanchor = 'auto', yshift = 0)) |> 
  # add web address
  layout(annotations = list(text = "www.dataviz-gwu.rocks", 
                            font = list(size = 10, color = 'grey'), showarrow = FALSE,
                            xref = 'paper', x = .5, xanchor = 'center', xshift = 0,
                            yref = 'paper', y = 1, yanchor = 'top', yshift = 0))
```


Notice that `plotly` has some fairly unique syntax for the `layout()` function. It helps to [read the documentation](https://plotly.com/python/reference/layout/annotations/#layout-annotations) but also to search around on google and Stack Overflow. Each annotation needs to be inputted in a list format. The first time in the list is the text you want to include in the annotation. From there you can include multiple arguments to specify the font size and location of the annotation.  
