---
title: Merging and Summarizing Data
date: today
date-format: long
footer: "[DataViz 2102 Website](https://dataviz-gwu.rocks)"
logo: images/dataviz-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## Merging Data Frames

- Often we have data from two different sources
- Results in two data frames
- How to make them one so we can analyze? 
- Key questions
  - What is the unit of analysis? 
  - What is/are the corresponding identifier variables?
  - Are the identifier variables in common? 
  - Or do they have to be added/transformed to match?
  
## Merging WB and V-Dem Data

- These are both time-series, country-level data
- Need to merge by country-year
- Year is easy
- But there are many different country codes
- Can use `countrycode` package to assign country codes

## `countrycode` Example

```{r}
#| label: countrycode
#| eval: false
#| echo: true

# Load countrycode
library(countrycode)

# Create new iso3c variable
democracy <- democracy |>    
  mutate(iso3c = countrycode(sourcevar = vdem_ctry_id, # what we are converting
        origin = "vdem",         # we are converting from vdem
        destination = "wb"))  |> # and converting to the WB iso3c code 
  relocate(iso3c, .after = vdem_ctry_id) # move iso3c 

# View the data
glimpse(democracy)
```

## Try it Yourself

- Using your democracy data frame from the last lesson
- Use `mutate()` and `countrycode()` to add iso3c country codes
- Use `relocate` to move your iso3c code to the "front" of your data frame (optional)

## Types of Joins in `dplyr`

- Mutating versus filtering joins
- Four types of mutating joins
  - `inner_join()`
  - `full_join()`
  - `left_join()`
  - `right_join()`
- For the most part we will use `left_join()`

## `left_join()` Example

```{r}
#| label: left_join()
#| eval: false
#| echo: true

# Load readr
library(readr)

# Perform left join using common iso3c variable and year
dem_women <- left_join(democracy, women_emp, by = c("iso3c", "year")) |> 
  rename(country = country.x) |> # rename country.x
  select(!country.y)             # crop country.y

# Save as .csv for future use
write_csv(dem_women, "data/dem_women.csv")

# View the data
glimpse(dem_women)  
```

## Try it Yourself

- Take your V-Dem data frame and your World Bank data frame
- Using `left_join()` to merge on country code and year
- Along the way, use `rename()` and `select()` to insure you have just one country name

## Group, Summarize and Arrange

- `group_by()`, `summarize()`, `arrange()`
- A very common sequence in data science:
  - Take an average or some other statistic for a group
  - Rank from high to low values of summary value
  
## Example

```{r}
#| label: summarize
#| eval: false
#| echo: true

# group_by(), summarize() and arrange()
dem_summary <- dem_women |> # save result as new object
  group_by(region)  |> # group dem_women data by region
  summarize(           # summarize following vars (by region)
    polyarchy = mean(polyarchy, na.rm = TRUE), # calculate mean, remove NAs
    gdp_pc = mean(gdp_pc, na.rm = TRUE), 
    flfp = mean(flfp, na.rm = TRUE), 
    women_rep = mean(women_rep, na.rm = TRUE)
  ) |> 
  arrange(desc(polyarchy)) # arrange in descending order by polyarchy score

# Save as .csv for future use
write_csv(dem_summary, "data/dem_summary.csv")

# View the data
glimpse(dem_summary)
```

## Try it Yourself {.scrollable}

- Try running a `group_by()`, `summarize()` and `arrange()` in your Quarto document
- Try changing the parameters to answer these questions: 

1. Try summarizing the data with a [different function](https://dplyr.tidyverse.org/reference/summarise.html#useful-functions) for one or more of the variables. 
  a. What is the median value of `polyarchy` for The West?
  b. What is the max value of `gdp_pc` for Eastern Europe?
  c. What is the standard deviation of `flfp` for Africa?
  d. What is the interquartile range of `women_rep` for the Middle East?
  
2. Now try grouping by country instead of region. 

  a. What is the median value of `polyarchy` for Sweden?
  b. What is the max value of `gdp_pc` New Zealand?
  c. What is the standard deviation of `flfp` for Spain?
  d. What is the interquartile range of `women_rep` for Germany?

3. Sort countries in descending order based on the mean value of `gdp_pc` (instead of the median value of `polyarchy`). Which country ranks first based on this sorting?

4. Now try sorting countries in ascending order based on the median value of `women_rep` (hint: delete "desc" from the `arrange()` call). Which country ranks at the "top" of the list?
