---
title: Downloading Data from an API
date: today
date-format: long
footer: "[DataViz 2102 Website](https://dataviz-gwu.rocks)"
logo: images/dataviz-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

## APIs

- API stands for "Application Programming Interface"
- Way for two computers to talk to each other
- In our case, we will use APIs to download social science data
- APIs are accessed through packages in R
- Sometimes there can be more than one package for an API
- Usually easier than reading in data from messy flat file

## This Lesson

- Access two APIs with R packages
- World Bank data through `wbstats`
  - There is another package called `WDI`
- Varieties of Democracy (V-Dem) through `vdemdata`
  - This is the only package for V-Dem
  - Just downloads all the data
  - So we have to use `dplyr` functions like `fliter()` and `select()`

## Searching for WB Indicators

```{r}
#| label: search_wb
#| eval: false
#| echo: true

flfp_indicators <- wb_search("female labor force") # store the list of indicators

print(flfp_indicators, n=26) # view the indicators
```
  
## Try it Yourself

- Start a new Quarto document
- Start a new code chunk
- Load the `wbstats` package
- Use `wb_search()` to find some indicators you are interested in

## `wbstats` Example

```{r}
#| label: download_wb
#| eval: false
#| echo: true

# Load packages
library(wbstats) # for downloading WB data
library(dplyr) # for selecting, renaming and mutating
library(janitor) # for rounding

# Store the list of indicators in an object
indicators <- c("flfp" = "SL.TLF.CACT.FE.ZS", "women_rep" = "SG.GEN.PARL.ZS") 

# Download the data  
women_emp <- wb_data(indicators, mrv = 50) |> # download data for last 50 yrs
  select(!iso2c) |> # drop the iso2c code which we won't be using
  rename(year = date) |> # rename date to year 
  mutate(
    flfp = round_to_fraction(flfp, denominator = 100), # round to nearest 100th
    women_rep = round_to_fraction(women_rep, denominator = 100) 
  )

# View the data
glimpse(women_emp) 
```

## Try it Yourself

- Using the variables you identified in your search, download the data
- Think about what elements of the code you need to change
- Hint: the indicators are going to change!

## Downloading V-Dem Data

- The `vdem` function from `vdemdata` just downloads the data
- There are no parameters for selecting indicators or date ranges
- So we have to do more work with `dplyr` functions
  - `filter()` for years
  - `select()` for variables
- We are also going to use `case_match()` to add 

## `vdemdata` Example

```{r}
#| label: vdemdata
#| eval: false
#| echo: true

# Load packages
library(vdemdata) # to download V-Dem data

# Download the data
democracy <- vdem |> # download the V-Dem dataset
  filter(year >= 1990)  |> # filter out years less than 1990
  select(                  # select (and rename) these variables
    country = country_name,     # the name before the = sign is the new name  
    vdem_ctry_id = country_id,  # the name after the = sign is the old name
    year, 
    polyarchy = v2x_polyarchy, 
    gdp_pc = e_gdppc, 
    region = e_regionpol_6C
    ) |>
  mutate(
    region = case_match(region, # replace the values in region with country names
                     1 ~ "Eastern Europe", 
                     2 ~ "Latin America",  
                     3 ~ "Middle East",   
                     4 ~ "Africa", 
                     5 ~ "The West", 
                     6 ~ "Asia")
                    # number on the left of the ~ is the V-Dem region code
                    # we are changing the number to the country name on the right
                    # of the equals sign
  )

# View the data
glimpse(democracy)
```

## Try it Yourself

- Go to the V-Dem [Codebook](https://www.v-dem.net/documents/24/codebook_v13.pdf) 
- Select a democracy indicator from Part III and one other indicator
- Note the indicator code (e.g. "v2x_polyarchy" for the polyarchy score)
- In your Quarto notes document, create a new code chunk
- Use the codes for your selected indicators to download the data
