---
title: Line Charts and Scatter Plots
date: today
date-format: long
footer: "[DataViz 6502 Website](https://dataviz-esia.training)"
logo: images/dataviz-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(vdemdata)
library(plotly)
library(WDI)

theme_set(theme_minimal())
```

```{r}
#| label: vdem_setup
#| include: false

polyarchy_annual <- vdem %>%
  group_by(year) %>%
  summarize(
    polyarchy = mean(v2x_polyarchy, na.rm = TRUE),
    gdp_pc = mean(e_gdppc, na.rm = TRUE)*1000
  ) %>%
  mutate(
    wave = case_when(year <= 1827 ~ "pre 1st wave",
                       year > 1827 & year <= 1926 ~ "1st wave", 
                       year > 1926 & year <= 1945 ~ "1st rev. wave",
                       year > 1945 & year <= 1962 ~ "2nd wave", 
                       year > 1962 & year <= 1973 ~ "2nd rev. wave", 
                       year > 1973 & year <= 2010 ~ "3rd wave", 
                       year > 2010 ~ "3rd rev. wave") %>%
      factor(levels = c("pre 1st wave", "1st wave", "1st rev. wave",
                        "2nd wave", "2nd rev. wave", "3rd wave", "3rd rev. wave"))
  ) %>%
  drop_na()
```

## Huntington's Three Waves

<br>

```{r}
#| label: huntington
region_plot <- ggplot(polyarchy_annual, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point(aes(color = wave)) + 
  geom_smooth(method = "lm", se = FALSE, size = .75) + 
  scale_x_log10(label = scales::label_dollar()) +
  aes(label = year) +
  labs(x= "GDP Per Capita", y = "Polyarchy Score",
       title = "Democracy, development and Huntington's three waves") +
  scale_color_discrete(name = "Wave")

ggplotly(region_plot, tooltip = c("year", "polyarchy")) %>%   
  # add source
  layout(annotations = list(text = "Source: The V-Dem Institute, Varities of Democracy Dataset",  
                            font = list(size = 10), showarrow = FALSE,
                            xref = 'paper', x = 1, xanchor = 'right', xshift = 100,
                            yref = 'paper', y = -.1, yanchor = 'auto', yshift = 0)) %>%
  # add web address
  layout(annotations = list(text = "www.psc2339.com", 
                            font = list(size = 10, color = 'grey'), showarrow = FALSE,
                            xref = 'paper', x = .5, xanchor = 'center', xshift = 0,
                            yref = 'paper', y = 1, yanchor = 'top', yshift = 0))
```

## Setup

```{r}
#| label: dem_waves_setup
#| echo: true
#| output: false

library(vdemdata)
library(tidyverse)

dem_waves_ctrs <- vdem |>
  select(
    country = country_name,     
    year, 
    polyarchy = v2x_polyarchy, 
  ) |>
  filter( 
    country %in% c("United States of America", # select countries in this list
                   "Japan", 
                   "Portugal")
    )

write_csv(dem_waves_ctrs, "data/dem_waves_ctrs.csv")
```

```{r}
#| label: exercise1_timer
countdown::countdown(minutes = 2, 
                     id = "setup", 
                     #bottom = "10%", 
                     #right = "10%",
                     color_border = "#fff",
                     color_text = "#fff",
                     color_running_background = "#42affa",
                     color_running_text = "black",
                     color_finished_background = "#E5D19D",
                     color_finished_text = "#00264A")
```

## Line Chart

```{r}
#| label: geom_line0
#| code-fold: true

# in this ggplot() call, we add a third dimension for line color
ggplot(dem_waves_ctrs, aes(x = year, y = polyarchy, color = country)) +
  geom_line(linewidth = 1) + # our geom is a line with a width of 1
  labs(
    x = "Year", 
    y = "Polyarchy Score", 
    title = 'Democracy in countries representing three different "waves"', 
    caption = "Source: V-Dem Institute", 
    color = "Country" # make title of legend to upper case
  )
```

## Line Chart

<br>

Use `geom_line()`... 

<br>

```{r}
#| label: geom_line1
#| echo: true
#| output: false
#| code-line-numbers: "3"

# in this ggplot() call, we add a third dimension for line color
ggplot(dem_waves_ctrs, aes(x = year, y = polyarchy, color = country)) +
  geom_line(linewidth = 1) + # our geom is a line with a width of 1
  labs(
    x = "Year", 
    y = "Polyarchy Score", 
    title = 'Democracy in countries representing three different "waves"', 
    caption = "Source: V-Dem Institute", 
    color = "Country" # make title of legend to upper case
  )
```

## Line Chart

<br>

Add third dimension for line color...

<br>

```{r}
#| label: geom_line2
#| echo: true
#| output: false
#| code-line-numbers: "2"

# in this ggplot() call, we add a third dimension for line color
ggplot(dem_waves_ctrs, aes(x = year, y = polyarchy, color = country)) +
  geom_line(linewidth = 1) + # our geom is a line with a width of 1
  labs(
    x = "Year", 
    y = "Polyarchy Score", 
    title = 'Democracy in countries representing three different "waves"', 
    caption = "Source: V-Dem Institute", 
    color = "Country" # make title of legend to upper case
  )
```

## Line Chart

<br>

Modify the legend title...

<br>

```{r}
#| label: geom_line3
#| echo: true
#| output: false
#| code-line-numbers: "9"

# in this ggplot() call, we add a third dimension for line color
ggplot(dem_waves_ctrs, aes(x = year, y = polyarchy, color = country)) +
  geom_line(linewidth = 1) + # our geom is a line with a width of 1
  labs(
    x = "Year", 
    y = "Polyarchy Score", 
    title = 'Democracy in countries representing three different "waves"', 
    caption = "Source: V-Dem Institute", 
    color = "Country" # make title of legend to upper case
  )
```

## Your Turn!

<br>

- See table three of [this article](https://kurzman.unc.edu/wp-content/uploads/sites/1410/2011/06/Kurzman_Waves_of_Democratization.pdf)
- Select three countries to visualize
- Adjust setup code to filter data on those countries 
- Visualize with `geom_line()`

```{r}
#|label: dem_waves_time
countdown::countdown(minutes = 10, 
                     id = "dem_waves", 
                     bottom = "10%", 
                     right = "10%")
```

## Scatter Plot Setup

<br>

```{r}
#| label: scatter_plot_setup
#| echo: true
#| output: false

dem_summary_ctry <- read_csv("https://raw.githubusercontent.com/eteitelbaum/dataviz-fall-2023/main/modules/data/dem_women.csv") |>
  group_by(country, region) |> # group by country, keep region
  summarize(
    polyarchy = mean(polyarchy, na.rm = TRUE),
    gdp_pc = mean(gdp_pc, na.rm = TRUE), 
    flfp = mean(flfp, na.rm = TRUE), 
    women_rep = mean(women_rep, na.rm = TRUE)
  )
```

```{r}
#|label: scatter_setup_time
countdown::countdown(minutes = 5, 
                     id = "scatter_setup", 
                     bottom = "10%", 
                     right = "10%")
```

## Scatter Plot

```{r}
#| label: scatter0
#| code-fold: true

ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy, color = region, size = women_rep)) + 
  geom_point() + # use geom_point() for scatter plots
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region",
    size = "Women Reps"
    )
```

## Scatter Plot

<br>

Use `geom_point()`...

<br>

```{r}
#| label: scatter1
#| echo: true
#| output: false
#| code-line-numbers: "2"

ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy, color = region, size = women_rep)) + 
  geom_point() + # use geom_point() for scatter plots
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region",
    size = "Women Reps"
    )
```

## Scatter Plot

<br>

Four dimensions... 

<br>

```{r}
#| label: scatter2
#| echo: true
#| output: false
#| code-line-numbers: "1"

ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy, color = region, size = women_rep)) + 
  geom_point() + # use geom_point() for scatter plots
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region",
    size = "Women Reps"
    )
```

## Scatter Plot

<br>

Stretch axis on log scale and use scales package to adjust labels...

```{r}
#| label: scatter3
#| echo: true
#| output: false
#| code-line-numbers: "3"

ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy, color = region, size = women_rep)) + 
  geom_point() + # use geom_point() for scatter plots
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region",
    size = "Women Reps"
    )
```


## Scatter Plot

<br>

Change legend titles... 

<br>

```{r}
#| label: scatter4
#| echo: true
#| output: false
#| code-line-numbers: "9-10"

ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy, color = region, size = women_rep)) + 
  geom_point() + # use geom_point() for scatter plots
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region",
    size = "Women Reps"
    )
```

## Your Turn! 

- There are four variables in `dem_summary_ctry`
- Pick one related to women's empowerment
- Visualize it on the y-axis with `gdp_pc` or `polyarch` on the x-axis 
- Change labels and legend titles to match your visualization
- Interpret your plot

```{r}
#|label: dem_waves_time
countdown::countdown(minutes = 10, 
                     id = "women_emp")
```

## Add a Trend Line

<br>

```{r}
#| label: trend_line1

ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point(aes(color = region)) + 
  geom_smooth(method = "lm", linewidth = 1) + 
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region"
    )
```

## Add a Trend Line

<br>

```{r}
#| label: trend_line2
#| echo: true
#| output: false
#| code-line-numbers: "3"
ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point(aes(color = region)) + 
  geom_smooth(method = "lm", linewidth = 1) + 
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region"
    )
```

## Add a Trend Line

<br>

Taking out size and adding color to `geom_point()` call...

<br>

```{r}
#| label: trend_line3
#| echo: true
#| output: false
#| code-line-numbers: "1-2"
ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point(aes(color = region)) + 
  geom_smooth(method = "lm", linewidth = 1) + 
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region"
    )
```

## Add a Trend Line

<br>

Changing legend titles.

<br>

```{r}
#| label: trend_line4
#| echo: true
#| output: false
#| code-line-numbers: "5-11"
ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point(aes(color = region)) + 
  geom_smooth(method = "lm", linewidth = 1) + 
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute", 
    color = "Region"
    )
```

## Your Turn! 

<br>

- Add a trendline to your plot
- Change the labels accordingly
- Try using `method = "loess"` instead of a "lm"

```{r}
#|label: trend_line_time
countdown::countdown(minutes = 5, 
                     id = "trend", 
                     bottom = "10%", 
                     right = "10%")
```

## Facet Wrapping

```{r}
#| label: facet_wrap0

ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point() + 
  geom_smooth(method = "lm", linewidth = 1) + 
  facet_wrap(~ region) +
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute"
    )
```

## Facet Wrapping

<br>

Use `facet_wrap()` with `~` before variable you want to wrap on...

```{r}
#| label: facet_wrap1
#| echo: true
#| output: false
#| code-line-numbers: "4"
ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point() + 
  geom_smooth(method = "lm", linewidth = 1) + 
  facet_wrap(~ region) +
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute"
    )
```

## Facet Wrapping

<br>

What else changes? Back down to two dimensions...

<br>

```{r}
#| label: facet_wrap2
#| echo: true
#| output: false
#| code-line-numbers: "1-2"
ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point() + 
  geom_smooth(method = "lm", linewidth = 1) + 
  facet_wrap(~ region) +
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute"
    )
```

## Facet Wrapping

<br>

Don't forget to take the legend title out of the captions...

<br>

```{r}
#| label: facet_wrap3
#| echo: true
#| output: false
#| code-line-numbers: "6-11"
ggplot(dem_summary_ctry, aes(x = gdp_pc, y = polyarchy)) + 
  geom_point() + 
  geom_smooth(method = "lm", linewidth = 1) + 
  facet_wrap(~ region) +
  scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
  labs(
    x= "GDP per Capita", 
    y = "Polyarchy Score",
    title = "Wealth and democracy, 1990 - present", 
    caption = "Source: V-Dem Institute"
    )
```

## Your Turn!

<br>

- Facet wrap your scatter plot
- Use `scales = "free"` to fix the West

```{r}
#|label: facet_wrap_time
countdown::countdown(minutes = 5, 
                     id = "facet", 
                     bottom = "10%", 
                     right = "10%")
```

## Labeling Points

```{r}
#| label: geom_text1
dem_summary_ctry |> 
  filter(region == "Asia") |>
  ggplot(aes(x = gdp_pc, y = polyarchy)) + 
    geom_point() + 
    geom_text(aes(label = country), size = 2, vjust = 2) +
    geom_smooth(method = "lm", linewidth = 1) +
    scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
      labs(
        x= "GDP Per Capita", 
        y = "Polyarchy Score",
        title = "Wealth and democracy in Asia, 1990 - present", 
        caption = "Source: V-Dem Institute"
        )
```

## Labeling Points

<br>

Filter for Asia, add labels with `geom_text()`...

<br>

```{r}
#| label: geom_text2
#| echo: true
#| output: false
#| code-line-numbers: "1-5"
dem_summary_ctry |> 
  filter(region == "Asia") |>
  ggplot(aes(x = gdp_pc, y = polyarchy)) + 
    geom_point() + 
    geom_text(aes(label = country), size = 2, vjust = 2) +
    geom_smooth(method = "lm", linewidth = 1) +
    scale_x_log10(labels = scales::label_number(prefix = "$", suffix = "k")) +
      labs(
        x= "GDP Per Capita", 
        y = "Polyarchy Score",
        title = "Wealth and democracy in Asia, 1990 - present", 
        caption = "Source: V-Dem Institute"
        )
```

## Your Turn!

<br>

- Filter for Asia or another region
- Use `geom_text()` to add labels to your points
- Play with `size` and `vjust` paramters 

```{r}
#|label: label_time
countdown::countdown(minutes = 5, 
                     id = "labels", 
                     bottom = "10%", 
                     right = "10%")
```



